\stdpopsim: a community-supported toolkit for population genetic simulation

### 2019-12-05 popSimManu.tex
Pass through intro.

\newcommand{\stdpopsim}{\texttt{stdpopsim} }
\newcommand{\dadi}{\texttt{dadi} }
\newcommand{\MSMC}{\texttt{MSMC} }
\newcommand{\smcpp}{\texttt{smcpp} }
\newcommand{\stairwayplot}{\texttt{stairwayplot} }
\newcommand{\fastsimcoal}{\texttt{fastsimcoal} }

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data 
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to 
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe the PopSim consortium project, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

Benchmarking method performance population genetics is not straightforward.
Many population genetic inference methods often make different assumptions as
well as use different summaries of the genetic variation data, even when
attempting to address the same biological question. For example, there are
numerous methods to infer population size changes.
The \MSMC method (\cite{schiffels2014inferring}) allows for the
population to continuously change in size over time. Other popular methods, like
\dadi (\cite{gutenkunst2009inferring}), typically allow for a smaller number of size changes to occur at certain
points throughout the population's history. Further, the types of data that the
methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
spatial patterns of genetic variation along the sequence, \dadi  instead uses the
site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
sample, ignoring any information about the correlation structure (linkage
disequilibrium) among SNPs. Thus from the start these methods may not be 
expected to perform equally well in all scenarios. \adk{I think we are focussed 
a bit too much on benchmarking currently in the intro. Could be good to add more on efficiency
and reproducibility}

The degree to which modeling assumptions and using different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
suggested large ancient (>100 years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see 
\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Further it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
for to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed simulations studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies are often specifically devised to showcase the performance
of the novel method and may not adquately compare performance to competing 
methods. Moreover researchers are generally forced to ``reinvent the wheel''
with respect to simulations, and are forced to reimplement previously
inferred scenarios independently. This can lead to errors being incorporated
during the implementation of the simulations. Lastly, the degree to
which different methods are compared can be quite variable.

One reason why simulation studies do not compare the performance of many methods
is that these simulation studies are challenging to carry out. There are several
practical impediments to carrying out the ideal benchmarking simulation studies.
First, while there are many published population genetic models available, their
parameters are often not easily accessible, often presented in supplemental
tables. Implementing these parameters into usable commands for simulations can
be a time-consuming and error prone process. Further, often papers present many
different models, and different authors may choose to select distinct models,
making comparison difficult. While improvements in simulation software and ever
faster computers have partially ameliorating this concern, generating the
simulated datasets stil can still be a time and computing resource bottleneck.

For these reasons, we have generated a standardized, community-driven resource 
for simulation of published demographic models from a number of popular study systems.
This resource, which we call PopSim, significantly eases the development
burden of implementing realistic simulations for population genetic analysis. 
Briefly, PopSim currently is comprised of a catalog of four organism genomes: humans,
\emph{Drosophila}, \emph{Arabidopsis}, and \emph{E. coli}. Each of these
genomes is associated with a physical organization (e.g., a chromosome structure), 
one or more genetic maps, default population-level parameters (mutation rate,
generation time) and one or more published demographic histories. Through 
either a command line interface or a mature Python API, the user can specify what
organism, map, chromosome(s), and history they are interested in simulating, and are returned
simulation output from their chosen model. PopSim has been developed using an 
completely distributed open source model, with strong procedures in place 
to continue its growth and accuracy. Below we describe PopSim and give an
example of how benchmarking of methods for demographic inference might proceed.

>>>>

\newcommand{\stdpopsim}{\texttt{stdpopsim} }
\newcommand{\dadi}{$\partial a \partial i$ }
\newcommand{\MSMC}{\texttt{MSMC} }
\newcommand{\smcpp}{\texttt{smcpp} }
\newcommand{\stairwayplot}{\texttt{stairwayplot} }
\newcommand{\fastsimcoal}{\texttt{fastsimcoal} }

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \texttt{stdpopsim}, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

Benchmarking the performance of inference methods in population genetics is not straightforward.
Inference methods often make different assumptions as
well as use different summaries of the genetic variation data, even when
attempting to address the same biological question. For example, there are
numerous methods to infer population size changes.
The \MSMC method (\cite{schiffels2014inferring}) allows for the
population to continuously change in size over time. Other methods, such as
\dadi (\cite{gutenkunst2009inferring}), allow for a smaller number of size changes to occur at certain
points throughout the population's history. Further, the types of data that the
methods use also differs. While \MSMC uses one to four whole genome sequences and leverages the
spatial patterns of genetic variation along the sequence, \dadi  instead uses the
site frequency spectrum (SFS), which summarizes the frequencies of SNPs in the
sample, ignoring any information about the correlation structure (linkage
disequilibrium) among SNPs. Thus from the start these methods may not be
expected to perform equally well in all scenarios. \adk{I think we are focussed
a bit too much on benchmarking currently in the intro. Could be good to add more on efficiency
and reproducibility}

The degree to which modeling assumptions and using different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed, simulations studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies are often specifically devised to showcase the performance
of the novel method and may not adequately compare performance to competing
methods. Moreover researchers are generally forced to ``reinvent the wheel''
with respect to simulations, and are forced to reimplement previously
inferred scenarios independently. This can lead to errors being incorporated
during the implementation of the simulations. Lastly, the degree to
which different methods are compared can be quite variable.

One reason why simulation studies do not compare the performance of many methods
is that these simulation studies are challenging to carry out. There are several
practical impediments to carrying out the ideal benchmarking simulation studies.
First, while there are many published population genetic models available, their
parameters are often not easily accessible, usually presented in supplemental
tables. Translating these parameters into input for a given simulation engine can
be a time-consuming and error prone process. In addition, papers often present many
different models, and different authors may choose to select distinct models,
making comparison difficult. While improvements in simulation software and ever
faster computers have partially ameliorating this concern, generating the
simulated datasets can still be a time and computing resource bottleneck.

For these reasons, we have generated a standardized, community-driven resource
for simulating  published demographic models from a number of popular study systems.
This resource, which we call \texttt{stdpopsim}, makes running
realistic simulations for population genetic analysis a simple matter of
choosing pre-implemented models from a community maintained catalog.
The \stdpopsim catalog currently contains four organisms: humans,
\emph{D.\ melanogaster}, \emph{A. thaliana}, and \emph{E. coli}. For each
organism the catalog contains details on the physical organization (e.g., chromosome structure)
of their genomes, one or more genetic maps, default population-level parameters (mutation rate,
generation time) and one or more published demographic histories. Through
either a command line interface or a simple Python API, the user can specify what
organism, map, chromosome, and demographic history they are interested in simulating, and are returned
simulation output from their chosen model.
The software has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and accuracy. Below we describe the resource and give an
example of how it can be used to benchmark demographic inference methods.
